---
title: "Module 4  - R Code"
author: "Sam Dumble"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}

library(dplyr)
library(ggplot2)
library(broom)

quakes<-datasets::quakes %>%
          mutate(order=1:1000)
quakes$cluster<-ifelse(quakes$long<175,"West","East")



```

## Exploratory Plots & Summaries

Comparing the smooth curve vs. the straight line
```{r}

ggplot(data=quakes,aes(y=stations ,x=mag ))+
 geom_smooth(method="lm",se=FALSE)+
geom_smooth(se=FALSE,col="red")+
  geom_point()

```

```{r}
cor.test(quakes$mag,quakes$stations)

```

## Fitting the linear regression model


```{r}
mod0<-lm(stations~mag,data=quakes)
```

```{r}
mod0

summary(mod0)
```

## Exploring Residuals

```{r}
plot(mod0)
```


```{r}

quakes$resid<-resid(mod0)
quakes$std_resid<-rstandard(mod0)

ggplot(quakes,aes(x=mag,y=stations,col=resid))+
  geom_point()+
     geom_smooth(method="lm",se=FALSE,col="blue")+
    scale_color_gradient2(high="red",low="red")
```


### Adding in Clusters

```{r}
ggplot(data=quakes,aes(y=lat ,x=long,fill=cluster,size=stations ))+
  geom_point(shape=21,alpha=0.5)
```

```{r}
ggplot(data=quakes,aes(y=stations ,x=cluster,colour=cluster ))+
  geom_violin(alpha=0.5)+
  geom_boxplot(width=0.2)
```

```{r}
quakes %>%
  group_by(cluster) %>%
    summarise(mean(stations),sd(stations))
```
```{r}
lm(stations~cluster,quakes) %>% summary()

t.test(stations~cluster,quakes,var.equal=TRUE)
```

### Adding cluster to mag

```{r}
ggplot(data=quakes,aes(y=stations ,x=mag,colour=cluster ))+
  geom_point()+
 geom_smooth(method="lm",se=FALSE)
```


```{r}
mod1<-lm(stations~mag+cluster,data=quakes)
summary(mod1)
```

### Exploring apparent contradictory trends in cluster

```{r}
ggplot(data=quakes,aes(y=mag ,x=cluster,colour=cluster ))+
  geom_violin(alpha=0.5)+
  geom_boxplot(width=0.2)
```

```{r}
quakes %>%
  mutate(magnitude=cut(mag,breaks=c(3.9,4.45,4.95,99),labels=c("4-4.49","4.5-4.99","5+")))%>%
  group_by(cluster,magnitude) %>%
    summarise(mean(stations),sd(stations),n=n()) %>%
  mutate(percent=scales::percent(n/sum(n),0.1)) 
```
## Adding an interaction

```{r}
lm(stations~mag*cluster,data=quakes) %>% summary()
```

## Making Predictions

Using "augment()" from the broom library here as it returns results in a dataframe that is a bit cleaner and easier to use!

```{r}
predict_data<-data.frame(mag=4.75,cluster="West")

augment(mod1,newdata=predict_data,interval="confidence")

augment(mod1,newdata=predict_data,interval="prediction")

```
## Assumptions

Simple versions of the key common plots can be made through plot(model):

```{r}
plot(mod1)
```

Exploring the time effect:

```{r}
quakes %>%
  mutate(resid=resid(mod1),
         time_group=as.numeric(cut(order,breaks=seq(0,1000,by=50),include.lowest = TRUE)))%>%
ggplot(aes(x=time_group,
                   y=resid,fill=stations))+
   stat_summary(geom="line")+
    stat_summary()+
    geom_smooth(se=FALSE)+
    xlab("Time Group")
```

Making a histogram of residuals:

```{r}
quakes %>%
  mutate(resid=resid(mod1))%>%
           ggplot(aes(x=resid))+
geom_histogram()
```

